---
title: Wrapup Blocking
format:
  html: default
  pdf: default
---

Note, that this dataset has been acquired with DDA (data dependent acquisition).
It was searched with MaxQuant. 
The analysis starts from the peptides.txt file, which is a table in the output directory of MaxQuant with the data summarized at the peptide-level. 

The preprocessing steps are slightly different. 
The modeling concepts, however, are also applicable to DIA data. 


<details><summary> Movie clip </summary><p><iframe width="560" height="315"
src="https://www.youtube.com/embed/HE4KWfIOVp0"
frameborder="0"
style="display: block; margin: auto;"
allow="autoplay; encrypted-media" allowfullscreen></iframe></p></details>


# Import Data and Preprocessing 

```{r echo = FALSE}
source("R/knitr_setup.R")
```

## Data 
<details><summary> Click to see code  </summary><p>
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(limma)
library(QFeatures)
library(msqrob2)
library(plotly)
library(gridExtra)

library("BiocFileCache")
bfc <- BiocFileCache()
peptidesFile <- bfcrpath(bfc,"https://github.com/statOmics/msqrob2data/raw/refs/heads/main/dda/mouseTcell/peptidesRCB.txt")
peptidesFile2 <- bfcrpath(bfc,"https://github.com/statOmics/msqrob2data/raw/refs/heads/main/dda/mouseTcell/peptidesCRD.txt")
peptidesFile3 <- bfcrpath(bfc,"https://github.com/statOmics/msqrob2data/raw/refs/heads/main/dda/mouseTcell/peptides.txt")

ecols <- grep("Intensity\\.", names(read.delim(peptidesFile)))
qf <- readQFeatures(
  assayData = read.delim(peptidesFile),
  fnames = 1,
  quantCols = ecols,
  name = "peptides_raw")

ecols2 <- grep("Intensity\\.", names(read.delim(peptidesFile2)))
qf2 <- readQFeatures(
  assayData = read.delim(peptidesFile2),
  fnames = 1,
  quantCols = ecols2,
  name = "peptides_raw")

ecols3 <- grep("Intensity\\.", names(read.delim(peptidesFile3)))
qf3 <- readQFeatures(
  assayData = read.delim(peptidesFile3),
  fnames = 1,
  quantCols = ecols3,
  name = "peptides_raw")

### Design
colData(qf)$celltype <- substr(
  colnames(qf[["peptides_raw"]]),
  11,
  14) %>%
  unlist %>%  
  as.factor

colData(qf)$mouse <- qf[[1]] %>%
  colnames %>%
  strsplit(split="[.]")  %>%
  sapply(function(x) x[3]) %>%
  as.factor

colData(qf2)$celltype <- substr(
  colnames(qf2[["peptides_raw"]]),
  11,
  14) %>%
  unlist %>%  
  as.factor

colData(qf2)$mouse <- qf2[[1]] %>%
  colnames %>%
  strsplit(split="[.]")  %>%
  sapply(function(x) x[3]) %>%
  as.factor

colData(qf3)$celltype <- substr(
  colnames(qf3[["peptides_raw"]]),
  11,
  14) %>%
  unlist %>%  
  as.factor

colData(qf3)$mouse <- qf3[[1]] %>%
  colnames %>%
  strsplit(split="[.]")  %>%
  sapply(function(x) x[3]) %>%
  as.factor
```
</p></design>

## Preprocessing 


### Log-transform

<details><summary> Click to see code to log-transform the data </summary><p>

- Peptides with zero intensities are missing peptides and should be represented
with a `NA` value rather than `0`.

```{r}
qf <- zeroIsNA(qf, "peptides_raw") # convert 0 to NA

qf2 <- zeroIsNA(qf2, "peptides_raw") # convert 0 to NA

qf3 <- zeroIsNA(qf3, "peptides_raw") # convert 0 to NA
```

- Logtransform data with base 2

```{r}
qf <- logTransform(qf, base = 2, i = "peptides_raw", name = "peptides_log")

qf2 <- logTransform(qf2, base = 2, i = "peptides_raw", name = "peptides_log")

qf3 <- logTransform(qf3, base = 2, i = "peptides_raw", name = "peptides_log")
```
</p></details>


### Filtering
<details><summary> Click to see details on filtering </summary><p>

1. Handling overlapping protein groups

#Only use proteotypic proteins

```{r}
qf <- filterFeatures(
  qf, ~ Proteins != "" #& ## Remove failed protein inference
    #!grepl(";", Proteins)
  )

qf2 <-  filterFeatures(
  qf2, ~ Proteins != "" #& ## Remove failed protein inference
  #  !grepl(";", Proteins)
  )

qf3 <-  filterFeatures(
  qf3, ~ Proteins != "" # & ## Remove failed protein inference
    #!grepl(";", Proteins)
  )
```

2. Remove reverse sequences (decoys) and contaminants

We now remove the contaminants, peptides that map to decoy sequences, and proteins
which were only identified by peptides with modifications.

```{r}
qf <- filterFeatures(qf,~Reverse != "+")
qf <- filterFeatures(qf,~ Potential.contaminant != "+")

qf2 <- filterFeatures(qf2,~Reverse != "+")
qf2 <- filterFeatures(qf2,~ Potential.contaminant != "+")

qf3 <- filterFeatures(qf3,~Reverse != "+")
qf3 <- filterFeatures(qf3,~ Potential.contaminant != "+")
```
3. Drop peptides that were only identified in one sample

We keep peptides that were observed at least twice.

```{r}
nObs <- 3
n1 <- ncol(qf[["peptides_log"]])
n2 <- ncol(qf2[["peptides_log"]])
n3 <- ncol(qf3[["peptides_log"]])

(qf <- filterNA(qf, i = "peptides_log", pNA = (n1 - nObs) / n1))
(qf2 <- filterNA(qf2, i = "peptides_log", pNA = (n2 - nObs) / n2))
(qf3 <- filterNA(qf3, i = "peptides_log", pNA = (n3 - nObs) / n3))
```

</p></details>

## Normalization 

<details><summary> Click to see code to normalize the data </summary><p>
```{r}
qf <- sweep( #4. Subtract log2 norm factor column-by-column (MARGIN = 2)
  qf,
  MARGIN = 2, 
  STATS = nfLogMedianOfRatios(qf, i="peptides_log"), #1.
  i = "peptides_log",
  name = "peptides_norm"
)

qf2 <- sweep( #4. Subtract log2 norm factor column-by-column (MARGIN = 2)
  qf2,
  MARGIN = 2, 
  STATS = nfLogMedianOfRatios(qf2, i="peptides_log"), #1.
  i = "peptides_log",
  name = "peptides_norm"
)

qf3 <- sweep( #4. Subtract log2 norm factor column-by-column (MARGIN = 2)
  qf3,
  MARGIN = 2, 
  STATS = nfLogMedianOfRatios(qf3, i="peptides_log"), #1.
  i = "peptides_log",
  name = "peptides_norm"
)
```

</p></details>

## Summarization

<details><summary> Click to see code to summarize the data </summary><p>

```{r,warning=FALSE}
qf <- aggregateFeatures(qf,
 i = "peptides_norm",
 fcol = "Proteins",
 name = "proteins", 
 fun = function(X) iq::maxLFQ(X)$estimate
)


qf2 <- aggregateFeatures(qf2,
 i = "peptides_norm",
 fcol = "Proteins",
 name = "proteins", 
 fun = function(X) iq::maxLFQ(X)$estimate
 )

qf3 <- aggregateFeatures(qf3,
 i = "peptides_norm",
 fcol = "Proteins",
 name = "proteins",
 fun = function(X) iq::maxLFQ(X)$estimate
 )
```

</p></details>

# Data Exploration: what is impact of blocking? 

<details><summary> Click to see code </summary><p>
```{r}
levels(colData(qf3)$mouse) <- paste0("m",1:7)
mdsObj3 <- plotMDS(assay(qf3[["proteins"]]), plot = FALSE)
mdsOrig <- colData(qf3) %>%
  as.data.frame %>%
  mutate(mds1 = mdsObj3$x,
         mds2 = mdsObj3$y,
         lab = paste(mouse,celltype,sep="_")) %>%
  ggplot(aes(x = mds1, y = mds2, label = lab, color = celltype, group = mouse)) +
  geom_text(show.legend = FALSE) +
  geom_point(shape = 21) +
  geom_line(color = "black", linetype = "dashed") +
  xlab(
    paste0(
      mdsObj3$axislabel,
      " ",
      1, 
      " (",
      paste0(
        round(mdsObj3$var.explained[1] *100,0),
        "%"
        ),
      ")"
      )
    ) +
  ylab(
    paste0(
      mdsObj3$axislabel,
      " ",
      2, 
      " (",
      paste0(
        round(mdsObj3$var.explained[2] *100,0),
        "%"
        ),
      ")"
      )
    ) +
  ggtitle("Original (RCB)")+ 
  theme_minimal()

levels(colData(qf)$mouse) <- paste0("m",1:4)
mdsObj <- plotMDS(assay(qf[["proteins"]]), plot = FALSE)
mdsRCB <- colData(qf) %>%
  as.data.frame %>%
  mutate(mds1 = mdsObj$x,
         mds2 = mdsObj$y,
         lab = paste(mouse,celltype,sep="_")) %>%
  ggplot(aes(x = mds1, y = mds2, label = lab, color = celltype, group = mouse)) +
  geom_text(show.legend = FALSE) +
  geom_point(shape = 21) +
  geom_line(color = "black", linetype = "dashed") +
  xlab(
    paste0(
      mdsObj$axislabel,
      " ",
      1, 
      " (",
      paste0(
        round(mdsObj$var.explained[1] *100,0),
        "%"
        ),
      ")"
      )
    ) +
  ylab(
    paste0(
      mdsObj$axislabel,
      " ",
      2, 
      " (",
      paste0(
        round(mdsObj$var.explained[2] *100,0),
        "%"
        ),
      ")"
      )
    ) +
  ggtitle("Randomized Complete Block (RCB)")+ 
  theme_minimal()


levels(colData(qf2)$mouse) <- paste0("m",1:8)
mdsObj2 <- plotMDS(assay(qf2[["proteins"]]), plot = FALSE)
mdsCRD <- colData(qf2) %>%
  as.data.frame %>%
  mutate(mds1 = mdsObj2$x,
         mds2 = mdsObj2$y,
         lab = paste(mouse,celltype,sep="_")) %>%
  ggplot(aes(x = mds1, y = mds2, label = lab, color = celltype, group = mouse)) +
  geom_text(show.legend = FALSE) +
  geom_point(shape = 21) +
  xlab(
    paste0(
      mdsObj$axislabel,
      " ",
      1, 
      " (",
      paste0(
        round(mdsObj2$var.explained[1] *100,0),
        "%"
        ),
      ")"
      )
    ) +
  ylab(
    paste0(
      mdsObj$axislabel,
      " ",
      2, 
      " (",
      paste0(
        round(mdsObj2$var.explained[2] *100,0),
        "%"
        ),
      ")"
      )
    ) +
  ggtitle("Completely Randomized Design (CRD)") + 
  theme_minimal()
```
</p></details>
```{r}
mdsOrig
mdsRCB
mdsCRD
```

- We observe that the leading fold change is according to mouse
- In the second dimension we see a separation according to cell-type 
- With the Randomized Complete Block design (RCB) we can remove the mouse effect from the analysis!

# Modeling and inference

## RCB analysis
```{r warning=FALSE}
qf <- msqrob(
  object = qf,
  i = "proteins",
  formula = ~ celltype + mouse)
```

## RCB wrong analysis
```{r warning=FALSE}
qf <- msqrob(
  object = qf,
  i = "proteins",
  formula = ~ celltype, modelColumnName = "wrongModel")
```

## CRD analysis 
```{r warning = FALSE}
qf2 <- msqrob(
  object = qf2,
  i = "proteins",
  formula = ~ celltype)
```

## Inference 

```{r}
library(ExploreModelMatrix)
VisualizeDesign(colData(qf),~ celltype + mouse)$plotlist
VisualizeDesign(colData(qf2),~ celltype)$plotlist
```


```{r}
L <- makeContrast("celltypeTreg = 0", parameterNames = c("celltypeTreg"))
qf <- hypothesisTest(object = qf, i = "proteins", contrast = L)
qf <- hypothesisTest(object = qf, i = "proteins", contrast = L, modelColumn = "wrongModel", resultsColumnNamePrefix="wrong")
qf2 <- hypothesisTest(object = qf2, i = "proteins", contrast = L)
```

# Advantage of Blocking: comparison between designs

## Volcano plots

<details><summary> Click to see code </summary><p>
```{r}
volcanoRCB <- ggplot(
    rowData(qf[["proteins"]])$celltypeTreg,
    aes(x = logFC, y = -log10(pval), color = adjPval < 0.05)
) +
    geom_point(cex = 2.5) +
    scale_color_manual(values = alpha(c("black", "red"), 0.5)) +
    theme_minimal() +
    ggtitle(paste0("RCB: \n", 
                sum(rowData(qf[["proteins"]])$celltypeTreg$adjPval<0.05,na.rm=TRUE),
            " significant"))

volcanoRCBwrong <- ggplot(
    rowData(qf[["proteins"]])$wrongcelltypeTreg,
    aes(x = logFC, y = -log10(pval), color = adjPval < 0.05)
) +
    geom_point(cex = 2.5) +
    scale_color_manual(values = alpha(c("black", "red"), 0.5)) +
    theme_minimal() +
    ggtitle(paste0("RCB wrong: \n", 
                sum(rowData(qf[["proteins"]])$wrongcelltypeTreg$adjPval<0.05,na.rm=TRUE),
            " significant"))

volcanoCRD <- ggplot(
    rowData(qf2[["proteins"]])$celltypeTreg,
    aes(x = logFC, y = -log10(pval), color = adjPval < 0.05)
) +
    geom_point(cex = 2.5) +
    scale_color_manual(values = alpha(c("black", "red"), 0.5)) +
    theme_minimal() +
    ggtitle(paste0("CRD: \n", 
                sum(rowData(qf2[["proteins"]])$celltypeTreg$adjPval<0.05,na.rm=TRUE),
            " significant"))
```
</p></details>
  
```{r}
grid.arrange(volcanoRCB,volcanoCRD, volcanoRCBwrong,ncol=2)
```

## Anova table: Q7TPR4, Alpha-actinin-1

Disclaimer: the Anova analysis is only for didactical purposes. In practice we assess the hypotheses using msqrob2. 

- We illustrate the power gain of blocking using an Anova analysis on 1 protein. 

- Note, that msqrob2 will perform a similar analysis, but, it uses robust regression and it uses an empirical Bayes estimator for the variance.

```{r}
prot <- "Q7TPR4"
dataHlp <- colData(qf) %>% 
  as.data.frame %>%
  mutate(intensity=assay(qf[["proteins"]])[prot,],
         intensityCRD=assay(qf2[["proteins"]])[prot,])

  anova(lm(intensity~ celltype + mouse, dataHlp)) 
  anova(lm(intensity~ celltype,dataHlp))
  anova(lm(intensityCRD~ celltype,dataHlp))
```

## Comparison Empirical Bayes standard deviation in msqrob2 

<details><summary> Click to see code </summary><p>
```{r}
accessions <- rownames(qf[["proteins"]])[rownames(qf[["proteins"]])%in%rownames(qf2[["proteins"]])]
dat <- data.frame(
sigmaRBC = sapply(rowData(qf[["proteins"]])$msqrobModels[accessions], getSigmaPosterior),
sigmaRBCwrong = sapply(rowData(qf[["proteins"]])$wrongModel[accessions], getSigmaPosterior),
sigmaCRD <- sapply(rowData(qf2[["proteins"]])$msqrobModels[accessions], getSigmaPosterior)
)

 plotRBCvsWrong <- ggplot(data = dat, aes(sigmaRBC, sigmaRBCwrong)) +
    geom_point(alpha = 0.1, shape = 20) +
    scale_x_log10() +
    scale_y_log10() +
    geom_abline(intercept=0,slope=1) +
   theme_minimal()
plotCRDvsWrong <- ggplot(data = dat, aes(sigmaCRD, sigmaRBCwrong)) +
    geom_point(alpha = 0.1, shape = 20) +
    scale_x_log10() +
    scale_y_log10() +
    geom_abline(intercept=0,slope=1) +   theme_minimal()

plotRBCvsCRD <- ggplot(data = dat, aes(sigmaRBC, sigmaCRD)) +
    geom_point(alpha = 0.1, shape = 20) +
    scale_x_log10() +
    scale_y_log10() +
    geom_abline(intercept=0,slope=1) +    theme_minimal()
```
</p></details>

```{r}
grid.arrange(
  plotRBCvsWrong,
  plotCRDvsWrong,
  plotRBCvsCRD,
  nrow=2)
```

- We clearly observe that the standard deviation of the protein expression in the RCB is smaller for the majority of the proteins than that obtained with the CRD

- The standard deviation of the protein expression RCB where we perform a wrong analysis without considering the blocking factor according to mouse is much larger for the majority of the proteins than that obtained with the correct analysis.

- Indeed, when we ignore the blocking factor in the RCB design we do not remove the variability according to mouse from the analysis and the mouse effect is absorbed in the error term. The standard deviation then becomes very comparable to that observed in the completely randomised design where we could not remove the mouse effect from the analysis. 

- Why are some of the standard deviations for the RCB with the correct analysis larger than that of the RCB with the incorrect analysis that ignored the mouse blocking factor?

- Can you think of a reason why it would not be useful to block on a particular factor? 
