# Heart use case: a MaxQuant LFQ DDA dataset with a more complex design {#sec-heart}

## Introduction

In this chapter we show how to analyse LFQ data from an experiment
with a more complex design. The data are a small subset of the public
dataset PXD006675 on PRIDE.

Particularly, the proteomes of the atrium and ventriculum in the left
and the right heart region are profiled for 3 patients (identifiers 3,
4, and 8). Hence, the design consists of a factor tissue (atrium,
ventriculum), region (left, right) and block (patient 3,4, and 8).

Suppose that researchers are mainly interested in comparing the
ventricular to the atrial proteome. Particularly, they would like to
compare the left atrium to the left ventricle, the right atrium to the
right ventricle, the average ventricular vs atrial proteome and if
ventricular vs atrial proteome shifts differ between left and right
heart region.

## Load packages

First, we load the `msqrob2` package and additional packages for data
manipulation and visualisation.

```{r}
suppressPackageStartupMessages({
library("QFeatures")  
library("dplyr") 
library("tidyr")
library("ggplot2")
library("msqrob2")    
library("stringr")
library("ExploreModelMatrix")
library("kableExtra")
library("ComplexHeatmap")
library("scater")
library("patchwork")
})
```

We also configure the [parallelisation](#sec-parallel) framework.

```{r}
library("BiocParallel")
register(SerialParam())
```

## Load Data

### Getting the data


The data were searched with MaxQuant version 1.5.5.6 and are
deposited on the PRIDE repository
[PXD006675](https://www.ebi.ac.uk/pride/archive/projects/PXD006675).

In this chapter we use a small subset of the data that is available on
msqrob2data.

```{r}
library("BiocFileCache")
bfc <- BiocFileCache()
peptideFile <- bfcrpath(bfc, "https://github.com/statOmics/msqrob2data/raw/refs/heads/main/dda/heart/peptides.txt")
```

After downloading the files, we can load the peptide table, which is
in ["wide format"](#sec-peptide_table). Hence, each row represents a
single peptide and each quantification column (that starts with
`"Intensity"`) represents a single sample.

```{r, warning=FALSE, message=FALSE}
peptides <- data.table::fread(peptideFile, check.names = TRUE, integer64 = "double")
quantcols <- grep("Intensity[.]", names(peptides), value = TRUE)
```

```{r, echo=FALSE}
knitr::kable(head(peptides))
```

We now extract the [sample annotations](#sec-annotation_table). We
will build a table where each row in the annotation table contains
information for one sample (the table below shows the first 6 rows).
This information is extracted from the sample names.

```{r}
coldata <- data.frame(quantCols = quantcols) |>     
  mutate(sampleId = gsub("Intensity[.]", "", quantcols), 
         location  = substr(sampleId, 1, 1), # heart region left-right
         tissue  = substr(sampleId, 2, 2), # tissue Atrium-Ventriculum
         patient  = substr(sampleId, 3, 3)) # patient id
```

```{r, echo=FALSE}
knitr::kable(head(coldata))
```

### The `QFeatures` data class

We combine the two tables into a [`QFeatures` object](#sec-qfeatures).

```{r}
(pe <- readQFeatures(
  peptides, colData = coldata, fnames = "Sequence", name = "peptides"
))
```

We now have a `QFeatures` object with 1 set, containing `r
nrows(pe)[[1]]` rows (peptides) and `r ncols(pe)[[1]]` columns
(samples). 

## Data preprocessing

`msqrob2` relies on the `QFeatures` data structure, meaning that we
can directly make use of `QFeatures`' data preprocessing functionality
(see also the `QFeatures`
[documentation](https://rformassspectrometry.github.io/QFeatures/articles/Processing.html)).

### Encoding missing values

Peptides with zero intensities should be
[encoded](#sec-encode_missing) using `NA`.

```{r}
pe <- zeroIsNA(pe, "peptides")
```

We calculate how many non zero intensities we have per peptide and
this is often useful for filtering.

```{r heart_nNA_hist, fig.cap=" "}
naResults <- nNA(pe, "peptides")
data.frame(naResults$nNArows) |> 
  ggplot() +
  aes(x = nNA) +
  geom_histogram()
```

### PSM filtering

We filter features based on 3 criteria (see [PSM filtering]).

1. Remove failed protein inference

We remove peptides that could not be uniquely mapped to a protein.

```{r}
pe <- filterFeatures(pe,
  ~ Proteins != "" & ## Remove failed protein inference
    !grepl(";", Proteins)) ## Remove protein groups
```

2. Remove reverse sequences (decoys) and contaminants

We remove the contaminants and peptides that map to decoy sequences.
These features bear no information of interest and will reduce the
statistical power upon multiple test adjustment.

```{r}
pe <- filterFeatures(pe, ~ Reverse != "+" & Potential.contaminant != "+")
```

3. Remove highly missing peptides. 

We keep peptides that were observed at least 3 times out of the $n =
12$ samples, so we tolerate the following proportion of NAs:
$\text{pNA} = \frac{(n - 3)}{n} = 0.75$, so we keep peptides that are
observed in at least 25% of the samples.

```{r}
nObs <- 3
n <- ncol(pe[["peptides"]])
(pe <- filterNA(pe, i = "peptides", pNA = (n - nObs) / n))
```

We keep `r nrow(pe[["peptides"]])` peptides upon filtering.


### Identifications per sample 

```{r}
pe[,,"peptides"] |> 
  longForm(colvars = colnames(colData(pe)), 
           rowvars= c("Sequence", 
                      "Proteins")) |>
  data.frame() |>
  filter(!is.na(value)) |>
  group_by(location, tissue, sampleId) |>
  summarise(Precursors = length(unique(Sequence)),
            `Protein Groups` = length(unique(Proteins))) |> 
  pivot_longer(-(1:3),
               names_to = "Feature",
               values_to = "IDs") |> 
  ggplot(aes(x = sampleId, y = IDs, fill = interaction(location,tissue))) +
  geom_col() +
  #scale_fill_observable() +
  facet_wrap(~Feature,
             scales = "free_y") +
  labs(title = "Peptide and protein group identifications per sample",
       x = "Sample",
       y = "Identifications") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90))
```

There are rather large variations in the peptide and protein identifications 
across samples. Therefore we will normalise the data using the Median of Ratios 
method of DESeq2, which can correct for differences in overall loading across 
runs as well as for differences in sample composition. 

### Standard preprocessing workflow

We can now prepare the data for modelling. The workflow ensures the
data complies to `msqrob2`'s requirements:

1. Intensities are [log-transformed](#sec-log2).

```{r}
pe <- logTransform(pe, base = 2, i = "peptides", name = "peptides_log")
```

2. Normalisation with Median of Ratios method.

```{r}
pe <- sweep( #Subtract log2 norm factor column-by-column (MARGIN = 2)
  pe,
  MARGIN = 2,
  STATS = nfLogMedianOfRatios(pe,"peptides_log"),
  i = "peptides_log",
  name = "peptides_norm"
)
```


Upon the normalisation the density curves should be nicely centred. To
confirm this, we will plot the intensity distributions for each
biorepeat (patient). `longForm()` seamlessly combines the quantification
and annotation data into a table suitable for `ggplot2` visualisation.
We also subset the object with the data before and after normalisation.

```{r heart_after_norm, fig.cap=" "}
longForm(pe[, , c("peptides_log", "peptides_norm")], colvar = "patient") |> 
  ggplot() +
  aes(x = value, group = colname, color = patient) +
  geom_density() +
  facet_wrap(~ assay, scale = "free")
```

3. [Summarisation](#sec-summarisation) to protein level.

We use median summarisation to infer protein-level data from
peptide-level data: for each protein and sample, the protein-level
intensity is taken as the median of the log2 intensities of its
peptides.

```{r,warning=FALSE}
pe <- aggregateFeatures(
  pe, i = "peptides_norm", fcol = "Proteins", 
  fun = matrixStats::colMedians, 
  na.rm = TRUE, name = "proteins"
)
```

## Data exploration

We will explore the main sources of variation in the data using
[MDS](#sec_data_exploration).

```{r heart_mds, fig.cap=" "}
library("scater")
se <- getWithColData(pe, "proteins") |> 
  as("SingleCellExperiment") |> 
  runMDS(exprs_values = 1) 
plotMDS(se, colour_by = "tissue") +
  plotMDS(se, colour_by = "location") +
  plotMDS(se, colour_by = "patient")
```

Note, that the samples upon median summarisation show a clear
separation according to the tissue type in the first dimension and
according to location in the second dimension.

## Data modelling

The preprocessed data can now be modelled to answer biologically
relevant questions. Particularly, the protein abundance can differ
according to tissue type (A-V) and location (L-R). Moreover, the
effect of the tissue type can differ according to the location and
vice versa. Hence, there can be an interaction between tissue and
location.

The samples are also not independent as four biopsies (LA, RA, LV and
RV) were taken for each patient. Because the proteome is profiled for
each tissue x location combination within each patient, the design is
a randomised complete block (RCB) design.

RCB designs can be correctly analysed by incorporating the block
effect for patient either as a fixed or a random effect. The use of a
fixed patient effect is here also possible because the effect of each
factor combination can be estimated within block (patient).

Here, we choose to account for the patient effect using fixed effects
because mixed models are computationally more demanding and rely on
asymptotic inference (i.e.  statistical inference is only valid for
experiments with large sample sizes).

Now we have identified the sources of variation in the experiment that
we have to account for (tissue, location and patient id), we can
define a model.

```{r warning=FALSE}
model <- ~ location*tissue + ## (1) fixed effects: main effects for location and tissue type, and a tissue x location interaction
  patient  ## (2) fixed block effect for patient
```

### Estimate the model

We estimate the model with `msqrob()`. Recall that
variables defined in `model` are automatically retrieved from the
`colData` (i.e. `"tissue"`, `"location"`, and `"patient"`). 

```{r, cache=TRUE, warning = FALSE, message = FALSE}
pe <- msqrob(
  pe, i = "proteins", formula = model, robust = TRUE
)
```

## Statistical inference

Once the models are estimated, we can start answering biological
questions by performing [Statistical inference]. We must translate the
biological questions into a statistical hypotheses:

 1. Is there an effect of tissue type (V-A) in the left heart region?
 2. Is there an effect of tissue type (V-A) in the right heart region?
 3. Is there on average an effect of tissue type in the heart. 
 4. Does the effect of tissue type (V-A) differ according to the heart region (L-R)? 
 
In other words, we must translate these questions in a linear
combination of the model parameters, also referred to as a contrast.
To aid defining contrasts, we will visualise the experimental design
using the `ExploreModelMatrix` package.

```{r heart_VisualizeDesign, fig.cap=" "}
library("ExploreModelMatrix")
vd <- VisualizeDesign(
    sampleData =  colData(pe),
    designFormula = ~ location*tissue + patient,
    textSizeFitted = 4
)
vd$plotlist
```

### Research question 1: is there an effect of tissue in the left heart region? 

From the plot we can see that the average log2 intensity for patient 3
in the left ventriculum equals `(Intercept) + tissueV`:

$$
\mu^L_{V,3} = \beta_0 + \beta_V
$$
and for the left atrium `(Intercept)`:

$$
\mu^L_{A,3} = \beta_0 
$$
So the average $\log_2 FC$ between atrium and ventriculum for patient
3 equals to parameter `tissueV`

$$
\log_2 FC_{V-A}^L = \mu^L_{V,3} -\mu^L_{A,3} = \beta_V
$$
The same can be seen for patient 4: 

$$
\log_2 FC_{V-A}^L= \mu^L_{V,4} -\mu^L_{A,4} = \beta_0 + \beta_V + \beta_4 - (\beta_0 + \beta_4) = \beta_V
$$
So the parameter `tissueV` has the interpretation of the average
$\log_2 FC$ between ventriculum and atrium after correction for the
patient effect, which quantifies the effect size for the first
research hypothesis.

### Research question 2: is there an effect of tissue in the right heart region? 

When we use the same rationale for the right heart region, we can see
that the average $\log_2 FC$ between atrium and ventriculum upon
correction for the patient effect equals `tissueV +
locationR:tissueV`. So, it consists of the main effect for tissue and
the location x tissue interaction.

We will illustrate this here for patient 4: 

$$
\begin{array}{rcl}
\log_2 FC_{V-A}^R& =&\mu^R_{V,4} -\mu^R_{A,4} \\
&=& \beta_0 + 
\beta_R + \beta_V + \beta_{R:V} + \beta_4 - (\beta_0 + 
\beta_R + \beta_4) \\
&=& \beta_V + \beta_{R:V}
\end{array}
$$

### Research question 3: is there an effect of tissue on average in the heart? 

This research question can be quantified by averaging
the $\log_2$ fold change between Ventriculum and Atrium over the left
and right heart regions, which equals `tissueV +
0.5*locationR:tissueV`

$$
\begin{array}{rcl}
(\log_2 FC_{V-A}^L + \log_2 FC_{V-A}^R)/ 2 &=& (\beta_V + \beta_V + \beta_{R:V})/2 \\
&=& \beta_V + 0.5\times\beta_{R:V}
\end{array}
$$

### Research question 4: does the effect of tissue differ according to the heart region?

This research question can be quantified by calculating the difference
in the $\log_2$ fold change between Ventriculum and Atrium in the
right and left heart regions, which equals `locationR:tissueV`

$$
\begin{array}{rcl}
\log_2 FC_{V-A}^R- \log_2 FC_{V-A}^L &=& \beta_V + \beta_{R:V}-\beta_V  \\
&=& \beta_{R:V}
\end{array}
$$
### Setting up the contrasts 

We can set up the four contrasts:

1. We make the design matrix so that we can easily extract all 
   parameter names from the model 
2. We make the contrast matrix for the four contrasts 

```{r}
L <- makeContrast(
  c(
    "tissueV = 0",
    "tissueV + locationR:tissueV = 0",
    "tissueV + 0.5*locationR:tissueV = 0",
    "locationR:tissueV = 0"
  ),
  parameterNames = colnames(vd$designmatrix)
  )
```

We can now falsify the null hypothesis of each contrast:

```{r}
pe <- hypothesisTest(
  object = pe, i = "proteins", contrast = L
)
```


### Results tables for significant proteins

We first collect the results for all contrasts. 

```{r}
inferences <- 
  msqrobCollect(pe[["proteins"]], L) 
head(inferences)
```

Next, we will return tables of proteins for which the contrasts are significant at the 5% FDR level. 

1. We set the significance level 
2. We loop over the contrasts
3. We filter the significant results for the contrast from the table
4. We print the table 

```{r, results='asis'}
alpha <- 0.05 #1.
for (contrasti in colnames(L)) #2. 
{
  sigList <- inferences |> 
    filter(contrast == contrasti & adjPval < alpha) #3.
  cat("**Contrast:**", contrasti, "= 0 (", nrow(sigList), " significant proteins)\n\n") #4.
  print(kable(sigList, row.names = FALSE) |>
      kable_styling(full_width = FALSE) |>
      scroll_box(height = "250px")
      ) #4.
  cat("\n\n\n---\n\n") #4.
}
```

Note, that no significant interactions are returned. 
This is not unexpected. 
We do not have a lot of subjects in the study and the power to pick up interaction terms is typically lower than to pick up main effects.


### Volcanoplots 

A volcano plot is a common visualisation that provides an overview of
the hypothesis testing results, plotting the $-\log_{10}$ p-value^[Note, 
that upon this transformation a value of 1 represents a p-value of 0.1, 2 
a p-value of 0.01, 3 a p-value of 0.001, etc.] as a function of
the estimated log fold change. Volcano plots can be used to highlight
the most interesting proteins that have large fold changes and/or are
highly significant. We can use the table above directly to build a
volcano plot using `ggplot2` functionality. We also highlight which
proteins are UPS standards, known to be differentially abundant by
experimental design.

Experienced users can make the plot themselves. However, in msqrob2 we also 
provide the plotVolcano function that generates volcanoplots based on `msqrob2`
inference tables generated with the `hypothesisTest` function. 

Since the inference table contains multiple contrasts we use the `facet_wrap`
function to make a separate volcano plot for each contrast. 

```{r}
volcanoplots <- inferences |> 
  plotVolcano() +
  facet_wrap(~contrast)
volcanoplots
```

```{r brixenHeartVolcanoTissueV}
 inferences |> filter(contrast == "tissueV") |> plotVolcano() + ggtitle("V - A (left)")
```

### Heatmaps

We use a nominal FDR level of 0.05

```{r}
alpha <- 0.05
```

We conduct heatmaps for the significant proteins for each contrast.

1. We first extract the assay `proteins` along with its colData. 
2. We make an empty list `heatmaps` to store the plots.
3. We will loop over the column names of the contrast matrix `L`
4. We extract the names of the significant features for contrast `i`
5. We extract the quant data for the significant proteins
6. We extract annotation
7. We make the heatmap using the quants data. We do not cluster columns (samples) to keep them together according to the design.

We only produce heatmaps for the first three contrasts as no DA proteins were returned for the location x tissue interaction.

```{r}
heatmaps <- lapply(colnames(L)[1:3],
                   function(contrast, se, alpha)
                   {
                     sig <- rowData(se)[[contrast]] |> 
                       filter(adjPval < alpha) |> 
                       rownames()
                     quants <- t(scale(t(assay(se[sig,]))))
                     colnames(quants) <- paste0(se$location, se$tissue, se$patient)
                     annotations <- columnAnnotation(condition = paste0(se$location, se$tissue))
                     set.seed(1234) ## annotation colours are randomly generated by default
                     return(
                       Heatmap(
                       quants, 
                       name = "log2 intensity",
                       top_annotation = annotations, 
                       column_title = paste0(contrast, " = 0")
                       )
                     )
                   },
                   se = getWithColData(pe, "proteins"),
                   alpha = alpha)
heatmaps
```

## Conclusion

In this chapter, we illustrated the analysis of a label-free
proteomics data set with biological replication. We followed the
workflow described in the previous chapters with minimal changes.

The experiment presented in this chapter has a complex design and
is an excellent illustration of how to model data with main effects, interactions and block effects.
We could investigate:

1. The difference in protein abundance between the atrium and 
   ventriculum in the **left** heart compartment.
2. The difference in protein abundance between the atrium and 
   ventriculum in the **right** heart compartment.
3. The difference in protein abundance between the atrium and 
   ventriculum, **averaged** over the right and left heart compartment.
4. The interaction, whether the difference between atrium and 
   ventriculum is affected by whether we focus on the left heart or 
   the right heart compartment.
   